---
title: "The Open Soil Index"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{the-open-soil-index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE}

  # load packages
  library(OBIC); library(data.table); library(ggplot2);require(patchwork)

  # load binnenveld data.table
  binnenveld <- as.data.table(OBIC::binnenveld)
```


## an introduction

To meet the demands of a growing population, agriculture continues to intensify, along with increasing and evolving impacts on crop growth, soil quality and environmental quality. There has been a great increase in world food production since the 1960s and an increase in per capita agricultural production, accompanied by a likewise increase in machinery and fertilizer use. Increased inputs of nitrogen and phosphorus to the soil however has also led to substantial negative impacts on biodiversity, drinking and surface water quality, and human health. Both the agronomic potential of cropping systems as well as the environmental impacts of agriculture are highly controlled by soil properties and its management. Soil health (or soil quality) refers to the capacity of soil to function as a living ecosystem that sustains plants, animals, and humans and support ecosystem services including agricultural production. Healthy soils are not just a growing medium for crops but they regulate and support a set of essential ecosystem services, such as water purification, carbon sequestration, nutrient cycling and they provide habitats for biodiversity. Improving and sustaining soil health is therefore key to sustainable crop production.

Many attempts have been made to develop indices to assess soil quality as well as related ecosystem services (reviewed by Bünemann et al., 2018; Rinot et al., 2019). To date, most of this knowledge has not been implemented in decision support tools assisting farmers for monitoring and evaluation of soil quality on farm and field level. Explicit evaluation of soil quality with respect to soil threats, soil functions and ecosystem services has rarely been implemented, and few approaches provide clear interpretation schemes of measured indicator values. The main reason for this lack of valorisation is that soil is a complex interplay of chemical, biological and physical processes and that the objectives for which the soil is used for has an enormous influence on the evaluation of soil quality. However, there is broad consensus that soil chemistry, soil structure, as well as soil biology need to be considered for a good interpretation of soil quality. 

In order to improve and maintain soil quality while supporting sustainable agronomic production as well as other soil ecosystems services, the Open Soil Index was initiatied in 2018, an open source framework for assessing and valorising soil quality and sustainable soil management. Experts from universities, consultancy companies, soil experts and farmer groups collaborated to bring together existing data from routine soil laboratories, smart data retrieval from satellites, insights from fundamental and applied research and smart in a transparent and scientifically underpinned soil assessment framework. The framework is highly influenced by the Soil Management Assessment Framework (Andrews & Carroll, 2001;Wienhold et al. 2004, 2009) and the proposed holistic, multivariate soil health framework from Rinot et al. (2019). 

## the Open Soil Index Framework
Soil quality is defined as the capacity of a soil to fulfil the desired soil functions under varying conditions for a combination of purposes (and services) such as food production, efficient nutrient cycling and preservation of biodiversity. Soil ecosystem services are the different ecosystem services that soils can provide, being categorised as production, support, regulation and information services. Soil functions are quantitative relationships evaluating how soil properties alone or in combination with each other contribute to the aforementioned ecosystem services. Soil indicators are the result of an evaluation scoring methodology transforming the output of soil functions to a qualitative category (low to high) or a numeric grade (1-10). Soil and field properties are characteristics of a soil (or field) that are used to quantify the soil functions. These characteristics may have been analysed in the laboratory as well as those resulting from the location of the field in the landscape. For example, the groundwater level, the variation in ground level, the slope, and the presence of drainage are soil properties, similar to the organic matter content, pH and clay content. Soil management includes all measures that land users or farmers can take to improve or adapt soil quality for a specific purpose. This has a direct influence on (measurable) soil properties. 

Within the boundary condition of context (i.e. continuation of crop rotation scheme in next decade) and objective (i.e. sustainable crop production), soil properties are quantified based on routinely available soil analyses, field properties and remote sensing data. The soil properties are in conjunction with each other used to quantify a number of soil functions. These functions can be clustered around the three relevant aspects of soil, namely i) chemistry and nutrient supply, ii) structure and root ability and iii) biology and disease resistance. Separately from these three aspects - each of which relies heavily on actual measurements - soil management is also evaluated. Furthermore, the soil's contribution to a sustainable living environment is first and foremost indirectly included as the following soil functions: the N buffering capacity for ground and surface water as well as for the potential to sequester carbon.
The algorithms to evaluate and assess the quality of the soil originates from field experiments and evidence from fundamental as well as applied research. 

<br />
<br />

# The Open Soil Index Calculator (OBIC)

The OBIC package makes it possible to evaluate Soil Health for a number of aggregated soil functions, evaluating a series of soil functions in relation to a "distance to target". 

## Data: Binnenveld
To explore the functionality of OBIC, we'll use the dataset `binnenveld`. The dataset contains soil properties from 11 agricultural fields in the neighborhood of Wageningen, with different soil texture and land use, and is documented in `?binnenveld`

For each field, the following properties are available:

* field properties: ID, year, landuse (B_LU_BRP) and agricultural region (B_AER_CBS)
* basic soil properties: soil type (B_SOILTYPE_AGR, B_HELP_WENR), risk for soil compaction (B_SC_WENR) and groundwater level (B_GWL_CLASS)
* chemical, physical and biological soil measurements done (all variables starting with "A_")
* measurements done via a Visual Soil Assessment Form (optional, all variables starting with "A_" and ending with "BCS"). For details, see the Assessment Form of the [Bodemconditiescore](http://mijnbodemconditie.nl/aan-de-slag/download/)
* soil management measures taken (optional, all variables starting with "M_")

```{r, showbinnenveld}

  dim(binnenveld)
  binnenveld[1]

```

In this example all optional input variables are set to the lowest score (for visuals soil assment) or to false (for soil management measures). Since soil quality can only be evaluated given its use, the crop rotation plan determines also which soil properties and which soil functions are relevant and how they are evaluated given a desired thresshold value.

Note that `binnenveld` is a data.table, a modern and fast variant of a data frame. It's particularly useful for large datasets because it only prints a selection of the dataset and calculations done on a data.table are extremely fast. You can learn more about data.tables at [this link](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) ; in particular you can convert data frmes to data.tables with as.data.table().

## Running OBIC for a single field
The main function to evaluate the quality of the soil is `obic_field()` as well as `obic_field_dt()`. Both are wrappers around the the following steps (where the last one requires a data.table as input and the first one not):

* Data preparation:
   * check format B_SC_WENR and B_GWL_CLASS
   * add default estimates for soil management measures given soil type and land use
* Estimate derivative soil properties
   * general supporting soil properties
   * soil chemical functions
   * soil physical functions
   * soil biological functions
   * soil environmental functions
   * soil management actions
   * visual soil assessment observations
* Estimate soil indicators (evaluating soil functions given a distance to target)
   * soil chemical functions
   * soil physical functions
   * soil biological functions
   * soil environmental functions
   * soil management actions
   * visual soil assessment score
* Aggregate soil indicators over the crop rotation plan
* Aggregate soil functions into more generic soil quality scores
   * for soil chemical, physical, biological and environmental soil quality (absolute)
   * for benchmarking (not yet implemented)

An example of both wrapper functions is illustrated below. The soil has an OBIC score of 0.81 indicating a relatively good quality to continue the current crop rotation plan. A value of one means that the soil quality is optimum for the continuation of the cropping plan whereas a value of zero means that there are substantial bottlenecks that need to be solved.


```{r}

  # select the relevant columns without management measures and data from Visual Soil Assessment
  cols <- colnames(binnenveld)[!grepl('_BCS$|^M_',colnames(binnenveld))]
  
  # select the first field 
  dt <- binnenveld[ID==1,mget(cols)]

  # run the obic_field with default management measures and no visual assessment data

  # test the obic field function via obic_field and give only the final score
  obic_field(dt$B_SOILTYPE_AGR,dt$B_GWL_CLASS,dt$B_SC_WENR,dt$B_HELP_WENR,dt$B_AER_CBS,
             dt$B_LU_BRP, dt$A_SOM_LOI, dt$A_SAND_MI, dt$A_SILT_MI, dt$A_CLAY_MI,dt$A_PH_CC,
             dt$A_CACO3_IF,dt$A_N_RT,dt$A_CN_FR,dt$A_COM_FR, dt$A_S_RT,dt$A_N_PMN,
             dt$A_P_AL, dt$A_P_CC, dt$A_P_WA,dt$A_CEC_CO,dt$A_CA_CO_PO, dt$A_MG_CO_PO, dt$A_K_CO_PO,
             dt$A_K_CC, dt$A_MG_CC, dt$A_MN_CC, dt$A_ZN_CC, dt$A_CU_CC,output = 'obic_score')
  
  # test the obic field function via obic_field_dt and give only the final score
  obic_field_dt(dt,output = 'obic_score')

```

When interested in the averaged indicator value or the averaged aggregated soil indicators, one can adjust the output argument of the wrapper functions.

```{r, results = FALSE}

  # run obic_field to retrieve indicators
  obic_field_dt(dt, output = 'indicators')

  # run obic_field to retrieve aggregated scores 
  # for chemistry, biology, fysics, management and environment
  obic_field_dt(dt, output = 'scores')
  
  # the default option is to retrieve all output
  obic_field_dt(dt, output = 'all')

```


<br />
<br />

# The Soil Functions in the OBIC
To assess the soil quality of an agricultural field, soil properties are used to derive so-called soil functions. Theses soil functions represent distinct processes in soil supporting the sustainable crop development. In this section the available soil functions are presented and illustrated.

## General supporting soil properties
Since the quality of a soil is evaluated given its crop rotation plan, soil type and geohydrological conditions, a few general derivative soil properties are estimated. These include the calculation of bulk density, rooting depth, organic carbon stocks, the age of grassland and the fraction that a crop contributes to the crop rotation plan. These general properties are explained below.

### Estimate bulk density
Soil Bulk density is estimated via a pedotransferfunction using 'calc_bulk_density' currently used in Dutch fertilizer recommendation systems. It requires as input the agronomic soil type, the soil organic matter content and the clay content of the topsoil analyzed. For details see `?calc_bulk_density`. Generally, the bulk density declines with the soil organic matter level in the soil and is higher in sandy soils compared to clayey soils.

<br />

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE, echo=FALSE}

  # make a data.table for sandy and clay soil
  dt.test <- data.table(B_SOILTYPE_AGR = c(rep('dekzand',10),rep('rivierklei',10)),
                        A_SOM_LOI = c(seq(0.1,10,length.out = 10), seq(0.1,10,length.out = 10)),
                        A_CLAY_MI = c(rep(5,10),rep(25,10)))

  # estimate bulk density (D_BDS)
  dt.test[, D_BDS := calc_bulk_density(B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI)]
  
  # plot output
  ggplot(data = dt.test,
         aes(x = A_SOM_LOI, y = D_BDS, group = B_SOILTYPE_AGR, color = B_SOILTYPE_AGR)) + 
    geom_line() + geom_point(size=3) +  theme_bw() + 
    ylab('Bulk density (kg / m3)') + xlab('Soil organic matter content (%)') + 
    theme(legend.position = c(0.8,0.8)) + ggtitle('Estimate bulk density from SOM and clay content') 
  
```


### Estimate other general properties
Agricultural soils are frequently analyzed in routine agricultural laboratories. In the Netherlands the soil layer analyzed is the top 25cm of an arable soil and the first 10 cm of a grassland soil. To convert soil nutrient concentrations (mg / kg) to a hectare basis, one need to know the soil depth analyzed. The OBIC function for this is a rather simple one, `calc_root_depth` where the details can be found via `?calc_root_depth`. The only input required is the crop code.

The Carbon pool in soil (kg / ha) is subsequently calculated from the soil organic matter content, the bulk density and the sample depth via `calc_organic_carbon`. For details, see `?calc_organic_carbon`.

The age of grassland is calculated via the function `calc_grass_age` given the series of crop codes available per field. These crop codes have a temporal order with the most recent ones on top. For details, see `?calc_grass_age`.

Lastly, the relative contribution of a few crops to the full crop rotation plan over the last ten years can be estimated via the function `calc_rotation_fraction`. This function has as input the field-ID, the series of crop codes and the name of the crop for which one likes to know the relative fraction within a crop rotation plan. The following options are possible: starch, potato, sugarbeet, grass, mais, alfalfa, catchcrop, cereal, clover, nature, rapeseed, other, rustgewas, and rustgewasdiep. 

An example of these functions is illustrated below.

```{r, results = FALSE}

  # estimate soil bulk density
  dt[, D_BDS := calc_bulk_density(B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI)]

  # estimate soil sampling depth
  dt[, D_RD := calc_root_depth(dt$B_LU_BRP)]

  # estimate Carbon pool
  dt[, D_OC := calc_organic_carbon(A_SOM_LOI, D_BDS, D_RD)]
  
  # estimate age of grassland
  dt[,D_GA := calc_grass_age(ID, B_LU_BRP)]
  
  # estimate crop rotation fraction for sugar beet
  dt[, D_CP_SUGARBEET := calc_rotation_fraction(ID, B_LU_BRP, crop = "sugarbeet")]
```


### Calculate series of chemical soil functions
The open soil index quantifies and evaluates the capacity of the soil to supply nitrogen, phosphorus, potassium, magnesium, copper and zinc as well as the capacity of the soil to buffer cations and the pH. The different functions are illustrated and explained below.

**Nitrogen supplying capacity**
<br />

Nitrogen supply on grassland is derived from empirical datasets relating total N levels in soil to the N supply, depending on sampling depth, soil texture and the age of the grassland. For arable fields, the N supply is derived from a simple first order decomposition model, calibrated for Dutch circumstances and depending on soil texture, organic matter and total N content. 

To estimate the N supplying capacity of the soil, one can make use of the function `calc_nlv`. For more details, see `?calc_nlv`. 

<br />

```{r, results = FALSE}

  # estimate nitrogen supply (kg N / ha)
  dt[, D_NLV := calc_nlv(B_LU_BRP, B_SOILTYPE_AGR, A_N_RT, A_CN_FR, D_OC, D_BDS, D_GA)]

```

<br />

An example of this function is illustrated below for both an arable and a grassland field (left figure) as well for the relationship between total N and NLV for a grassland soil. In summary, the N supplying capacity of a soil increases with the total N present in the soil.

<br />

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # define an arable field with two landuses
  dt.test <- binnenveld[ID %in% c(1,3)]
  
  # estimate default properties needed to estimate NLV
  dt.test[, D_BDS := calc_bulk_density(B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI)]
  dt.test[, D_RD := calc_root_depth(B_LU_BRP)]
  dt.test[, D_OC := calc_organic_carbon(A_SOM_LOI, D_BDS, D_RD)]
  dt.test[, D_GA := calc_grass_age(ID, B_LU_BRP)]

  # estimate nitrogen supply (kg N / ha)
  dt.test[, D_NLV := calc_nlv(B_LU_BRP, B_SOILTYPE_AGR, A_N_RT, A_CN_FR, D_OC, D_BDS, D_GA)]
  
  # add group for figure
  dt.test[,luse := fifelse(ID==1,'arable','grasland')]
  
  # estimate averaged NLV for both soils
  dt.test1 <- dt.test[,list(D_NLV = mean(D_NLV)), by = 'luse']
  
   # plot output
  p1 <- ggplot(data = dt.test1,aes(y = D_NLV, x = luse, fill = luse)) + geom_col() +  theme_bw() + 
        ylab('N supplying capacity (kg N / ha)') + xlab('Landuse') + 
        theme(legend.position = c(0.2,0.8),
              plot.title = element_text(size=10),
              legend.text = element_text(size=10),
              axis.text = element_text(size=10)) + ggtitle('N supplying capacity')
     
  
  # estimate NLV for grassland soil over range of N-total levels in soil
  dt.test2 <- dt.test[ID==3]
  
  # overwrite N levels of the soil and the soil texture to sand with a range to illustrate impact of N total on NLV
  dt.test2[, A_N_RT := seq(100,5000,length.out = .N)]
  dt.test2[, B_SOILTYPE_AGR := 'dekzand']
  dt.test2[, D_GA := 5]
  dt.test2[, D_NLV := calc_nlv(B_LU_BRP, B_SOILTYPE_AGR, A_N_RT, A_CN_FR, D_OC, D_BDS, D_GA)]
    
  # plot output
  p2 <- ggplot(data = dt.test2,
               aes(x = A_N_RT, y = D_NLV)) + geom_line() + geom_point(size=3) +  theme_bw() + 
    theme(plot.title = element_text(size=10),
              legend.text = element_text(size=10),
              axis.text = element_text(size=10))+
    ylab('N supplying capacity (kg N / ha)') + xlab('Total Nitrogen content (mg / kg)') +
    ggtitle('Relationship NLV (grasland) and N total') 

  # plot side by side
  p1 + p2
  
```


The total N supply is evaluated via a parabolic scoring function with an optimum of 100 kg N / ha in arable fields and 140 kg N / ha in grassland fields.

<br />

**Phosphorus Availability Index**
<br />

The phosphorus supply is derived from one capacity and one intensity P pools in soil for grassland and maize fields, where the optimum P supply is derived from multiple field experiments across the Netherlands. For arable farming systems, the phosphorus supply is derived from one measured P pool reflecting the P status as well as controlling the crop response to P inputs. The availability is highly controlled by chemical sorption and desorption equilibria, and affected by the iron and aluminum oxides content of soils as well as the availability of oxygen. A phosphorus supply around 4.8 (unitless index, for maize and grassland fields) and 45 (mg P2O5 per liter, for arable fields)  is optimal for crop production.




