---
title: "The Open Soil Index"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{the-open-soil-index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE}

  # load packages
  library(OBIC); library(data.table); library(ggplot2);require(patchwork); library(ggpubr)

  # load binnenveld data.table
  binnenveld <- as.data.table(OBIC::binnenveld)
```


## an introduction

To meet the demands of a growing population, agriculture continues to intensify, along with increasing and evolving impacts on crop growth, soil quality and environmental quality. There has been a great increase in world food production since the 1960s and an increase in per capita agricultural production, accompanied by a likewise increase in machinery and fertilizer use. Increased inputs of nitrogen and phosphorus to the soil however has also led to substantial negative impacts on biodiversity, drinking and surface water quality, and human health. Both the agronomic potential of cropping systems as well as the environmental impacts of agriculture are highly controlled by soil properties and its management. Soil health (or soil quality) refers to the capacity of soil to function as a living ecosystem that sustains plants, animals, and humans and support ecosystem services including agricultural production. Healthy soils are not just a growing medium for crops but they regulate and support a set of essential ecosystem services, such as water purification, carbon sequestration, nutrient cycling and they provide habitats for biodiversity. Improving and sustaining soil health is therefore key to sustainable crop production.

Many attempts have been made to develop indices to assess soil quality as well as related ecosystem services (reviewed by Bünemann et al., 2018; Rinot et al., 2019). To date, most of this knowledge has not been implemented in decision support tools assisting farmers for monitoring and evaluation of soil quality on farm and field level. Explicit evaluation of soil quality with respect to soil threats, soil functions and ecosystem services has rarely been implemented, and few approaches provide clear interpretation schemes of measured indicator values. The main reason for this lack of valorisation, is that soil is a complex interplay of chemical, biological and physical processes and that the objectives for which the soil is used, have an enormous influence on the evaluation of soil quality. However, there is broad consensus that soil chemistry, soil structure, as well as soil biology, need to be considered for a good interpretation of soil quality. 

In order to improve and maintain soil quality while supporting sustainable agronomic production as well as other soil ecosystems services, the Open Soil Index was initiated in 2018, an open source framework for assessing and valorising soil quality and sustainable soil management. Experts from universities, consultancy companies, soil experts and farmer groups collaborated to bring together existing data from routine soil laboratories, smart data retrieval from satellites, insights from fundamental and applied research in a transparent and scientifically underpinned soil assessment framework. The framework is highly influenced by the Soil Management Assessment Framework (Andrews & Carroll, 2001;Wienhold et al. 2004, 2009) and the proposed holistic, multivariate soil health framework from Rinot et al. (2019). 

## the Open Soil Index Framework
Soil quality is defined as the capacity of a soil to fulfill the desired soil functions under varying conditions for a combination of purposes (and services) such as food production, efficient nutrient cycling and preservation of biodiversity. Soil ecosystem services are the different ecosystem services that soils can provide, being categorized as production, support, regulation and information services. Soil functions are quantitative relationships evaluating how soil properties alone or in combination with each other contribute to the aforementioned ecosystem services. Soil indicators are the result of an evaluation scoring methodology transforming the output of soil functions to a qualitative category (low to high) or a numeric grade (1-10). Soil and field properties are characteristics of a soil (or field) that are used to quantify the soil functions. These characteristics may have been analyzed in the laboratory as well as those resulting from the location of the field in the landscape. For example, the groundwater level, the variation in ground level, the slope, and the presence of drainage are soil properties, similar to the organic matter content, pH and clay content. Soil management includes all measures that land users or farmers can take to improve or adapt soil quality for a specific purpose. This has a direct influence on (measurable) soil properties. 

Within the boundary condition of context (i.e. continuation of crop rotation scheme in next decade) and objective (i.e. sustainable crop production), soil properties are quantified based on routinely available soil analyses, field properties and remote sensing data. The soil properties are in conjunction with each other used to quantify a number of soil functions. These functions can be clustered around the three relevant aspects of soil, namely i) chemistry and nutrient supply, ii) structure and root ability and iii) biology and disease resistance. Separately from these three aspects - each of which relies heavily on actual measurements - soil management is also evaluated. Furthermore, the soil's contribution to a sustainable living environment is first and foremost indirectly included as the following soil functions: the N buffering capacity for ground and surface water as well as for the potential to sequester carbon.
The algorithms to evaluate and assess the quality of the soil originates from field experiments and evidence from fundamental as well as applied research. 

<br />
<br />

# The Open Soil Index Calculator (OBIC)

The OBIC package makes it possible to evaluate Soil Health for a number of aggregated soil functions, evaluating a series of soil functions in relation to a "distance to target". 

## Data: Binnenveld
To explore the functionality of OBIC, we'll use the dataset `binnenveld`. The dataset contains soil properties from 11 agricultural fields in the neighborhood of Wageningen, with different soil texture and land use, and is documented in `?binnenveld`

For each field, the following properties are available:

* field properties: ID, year, landuse (B_LU_BRP) and agricultural region (B_AER_CBS)
* basic soil properties: soil type (B_SOILTYPE_AGR, B_HELP_WENR), risk for soil compaction (B_SC_WENR) and groundwater level (B_GWL_CLASS)
* chemical, physical and biological soil measurements done (all variables starting with "A_")
* measurements done via a Visual Soil Assessment Form (optional, all variables starting with "A_" and ending with "BCS"). For details, see the Assessment Form of the [Bodemconditiescore](http://mijnbodemconditie.nl/aan-de-slag/download/)
* soil management measures taken (optional, all variables starting with "M_")

```{r, showbinnenveld}

  dim(binnenveld)
  binnenveld[1]

```

In this example all optional input variables are set to the lowest score (for visuals soil assment) or to false (for soil management measures). Since soil quality can only be evaluated given its use, the crop rotation plan determines also which soil properties and which soil functions are relevant and how they are evaluated given a desired threshold value.

Note that `binnenveld` is a data.table, a modern and fast variant of a data frame. It's particularly useful for large datasets because it only prints a selection of the dataset and calculations done on a data.table are extremely fast. You can learn more about data.tables at [this link](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) ; in particular you can convert data frames to data.tables with as.data.table().

## Running OBIC for a single field
The main function to evaluate the quality of the soil is `obic_field()` as well as `obic_field_dt()`. Both are wrappers around the the following steps (where the latter requires a data.table as input and the former does not):

* Data preparation:
   * check format B_SC_WENR and B_GWL_CLASS
   * add default estimates for soil management measures given soil type and land use
* Estimate derivative soil properties
   * general supporting soil properties
   * soil chemical functions
   * soil physical functions
   * soil biological functions
   * soil environmental functions
   * soil management actions
   * visual soil assessment observations
* Estimate soil indicators (evaluating soil functions given a distance to target)
   * soil chemical functions
   * soil physical functions
   * soil biological functions
   * soil environmental functions
   * soil management actions
   * visual soil assessment score
* Aggregate soil indicators over the crop rotation plan
* Aggregate soil functions into more generic soil quality scores
   * for soil chemical, physical, biological and environmental soil quality (absolute)
   * for benchmarking (not yet implemented)

An example of both wrapper functions is illustrated below. The soil has an OBIC score of 0.81 indicating a relatively good quality to continue the current crop rotation plan. A value of one means that the soil quality is optimum for the continuation of the cropping plan whereas a value of zero means that there are substantial bottlenecks that need to be solved.


```{r}

  # select the relevant columns without management measures and data from Visual Soil Assessment
  cols <- colnames(binnenveld)[!grepl('_BCS$|^M_',colnames(binnenveld))]
  
  # select the first field 
  dt <- binnenveld[ID==1,mget(cols)]

  # run the obic_field with default management measures and no visual assessment data

  # test the obic field function via obic_field and give only the final score
  obic_field(dt$B_SOILTYPE_AGR,dt$B_GWL_CLASS,dt$B_SC_WENR,dt$B_HELP_WENR,dt$B_AER_CBS,
             dt$B_LU_BRP, dt$A_SOM_LOI, dt$A_SAND_MI, dt$A_SILT_MI, dt$A_CLAY_MI,dt$A_PH_CC,
             dt$A_CACO3_IF,dt$A_N_RT,dt$A_CN_FR,dt$A_COM_FR, dt$A_S_RT,dt$A_N_PMN,
             dt$A_P_AL, dt$A_P_CC, dt$A_P_WA,dt$A_CEC_CO,dt$A_CA_CO_PO, dt$A_MG_CO_PO, dt$A_K_CO_PO,
             dt$A_K_CC, dt$A_MG_CC, dt$A_MN_CC, dt$A_ZN_CC, dt$A_CU_CC,output = 'obic_score')
  
  # test the obic field function via obic_field_dt and give only the final score
  obic_field_dt(dt,output = 'obic_score')

```

When interested in the averaged indicator value or the averaged aggregated soil indicators, one can adjust the output argument of the wrapper functions.

```{r, results = FALSE}

  # run obic_field to retrieve indicators
  obic_field_dt(dt, output = 'indicators')

  # run obic_field to retrieve aggregated scores 
  # for chemistry, biology, fysics, management and environment
  obic_field_dt(dt, output = 'scores')
  
  # the default option is to retrieve all output
  obic_field_dt(dt, output = 'all')

```


<br />
<br />

# The Soil Functions in the OBIC
To assess the soil quality of an agricultural field, soil properties are used to derive so-called soil functions. Theses soil functions represent distinct processes in soil supporting the sustainable crop development. In this section the available soil functions are presented and illustrated.

## General supporting soil properties
Since the quality of a soil is evaluated given its crop rotation plan, soil type and geohydrological conditions, a few general derivative soil properties are estimated. These include the calculation of bulk density, rooting depth, organic carbon stocks, the age of grassland and the fraction that a crop contributes to the crop rotation plan. These general properties are explained below.

### Estimate bulk density
Soil Bulk density is estimated via a pedotransferfunction using 'calc_bulk_density' currently used in Dutch fertilizer recommendation systems. It requires as input the agronomic soil type, the soil organic matter content and the clay content of the topsoil analyzed. For details see `?calc_bulk_density`. Generally, the bulk density declines with the soil organic matter level in the soil and is higher in sandy soils compared to clayey soils.

<br />

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE, echo=FALSE}

  # make a data.table for sandy and clay soil
  dt.test <- data.table(B_SOILTYPE_AGR = c(rep('dekzand',10),rep('rivierklei',10)),
                        A_SOM_LOI = c(seq(0.1,10,length.out = 10), seq(0.1,10,length.out = 10)),
                        A_CLAY_MI = c(rep(5,10),rep(25,10)))

  # estimate bulk density (D_BDS)
  dt.test[, D_BDS := calc_bulk_density(B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI)]
  
  # plot output
  ggplot(data = dt.test,
         aes(x = A_SOM_LOI, y = D_BDS, group = B_SOILTYPE_AGR, color = B_SOILTYPE_AGR)) + 
    geom_line() + geom_point(size=3) +  theme_bw() + 
    ylab('Bulk density (kg / m3)') + xlab('Soil organic matter content (%)') + 
    theme(legend.position = c(0.8,0.8)) + ggtitle('Estimate bulk density from SOM and clay content') 
  
```


### Estimate other general properties
Agricultural soils are frequently analyzed in routine agricultural laboratories. In the Netherlands the soil layer analyzed, is the top 25cm of an arable soil and the first 10 cm of a grassland soil. To convert soil nutrient concentrations (mg / kg) to a hectare basis, one need to know the soil depth analyzed. The OBIC function for this is a rather simple one, `calc_root_depth` where the details can be found via `?calc_root_depth`. The only input required is the crop code.

The Carbon pool in soil (kg / ha) is subsequently calculated from the soil organic matter content, the bulk density and the sample depth via `calc_organic_carbon`. For details, see `?calc_organic_carbon`.

The age of grassland is calculated via the function `calc_grass_age` given the series of crop codes available per field. These crop codes have a temporal order with the most recent ones on top. For details, see `?calc_grass_age`.

Lastly, the relative contribution of a few crops to the full crop rotation plan over the last ten years can be estimated via the function `calc_rotation_fraction`. This function has as input the field-ID, the series of crop codes and the name of the crop for which one likes to know the relative fraction within a crop rotation plan. The following options are possible: starch, potato, sugarbeet, grass, mais, alfalfa, catchcrop, cereal, clover, nature, rapeseed, other, rustgewas, and rustgewasdiep. 

An example of these functions is illustrated below.

```{r, results = FALSE}

  # estimate soil bulk density
  dt[, D_BDS := calc_bulk_density(B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI)]

  # estimate soil sampling depth
  dt[, D_RD := calc_root_depth(dt$B_LU_BRP)]

  # estimate Carbon pool
  dt[, D_OC := calc_organic_carbon(A_SOM_LOI, D_BDS, D_RD)]
  
  # estimate age of grassland
  dt[,D_GA := calc_grass_age(ID, B_LU_BRP)]
  
  # estimate crop rotation fraction for sugar beet
  dt[, D_CP_SUGARBEET := calc_rotation_fraction(ID, B_LU_BRP, crop = "sugarbeet")]
```


## Calculate series of chemical soil functions
The open soil index quantifies and evaluates the capacity of the soil to supply nitrogen, phosphorus, potassium, magnesium, copper and zinc as well as the capacity of the soil to buffer cations and the pH. The different functions are illustrated and explained below.

### Nitrogen supplying capacity
Nitrogen supply on grassland is derived from empirical datasets relating total N levels in soil to the N supply, depending on sampling depth, soil texture and the age of the grassland. For arable fields, the N supply is derived from a simple first order decomposition model, calibrated for Dutch circumstances and depending on soil texture, organic matter and total N content. 

To estimate the N supplying capacity of the soil, one can make use of the function `calc_nlv`. For more details, see `?calc_nlv`. 


```{r, results = FALSE}

  # estimate nitrogen supply (kg N / ha)
  dt[, D_NLV := calc_nlv(B_LU_BRP, B_SOILTYPE_AGR, A_N_RT, A_CN_FR, D_OC, D_BDS, D_GA)]

```


An example of this function is illustrated below for both an arable and a grassland field (left figure) as well for the relationship between total N and NLV for a grassland soil. In summary, the N supplying capacity of a soil increases with the total N present in the soil.

<br />

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # define an arable field with two landuses
  dt.test <- binnenveld[ID %in% c(1,3)]
  
  # estimate default properties needed to estimate NLV
  dt.test[, D_BDS := calc_bulk_density(B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI)]
  dt.test[, D_RD := calc_root_depth(B_LU_BRP)]
  dt.test[, D_OC := calc_organic_carbon(A_SOM_LOI, D_BDS, D_RD)]
  dt.test[, D_GA := calc_grass_age(ID, B_LU_BRP)]

  # estimate nitrogen supply (kg N / ha)
  dt.test[, D_NLV := calc_nlv(B_LU_BRP, B_SOILTYPE_AGR, A_N_RT, A_CN_FR, D_OC, D_BDS, D_GA)]
  
  # add group for figure
  dt.test[,luse := fifelse(ID==1,'arable','grasland')]
  
  # estimate averaged NLV for both soils
  dt.test1 <- dt.test[,list(D_NLV = mean(D_NLV)), by = 'luse']
  
   # plot output
  p1 <- ggplot(data = dt.test1,aes(y = D_NLV, x = luse, fill = luse)) + geom_col() +  theme_bw() + 
        ylab('N supplying capacity (kg N / ha)') + xlab('Landuse') + 
        theme(legend.position = c(0.2,0.8),
              plot.title = element_text(size=10),
              legend.text = element_text(size=10),
              axis.text = element_text(size=10)) + ggtitle('N supplying capacity')
     
  
  # estimate NLV for grassland soil over range of N-total levels in soil
  dt.test2 <- dt.test[ID==3]
  
  # overwrite N levels of the soil and the soil texture to sand with a range to illustrate impact of N total on NLV
  dt.test2[, A_N_RT := seq(100,5000,length.out = .N)]
  dt.test2[, B_SOILTYPE_AGR := 'dekzand']
  dt.test2[, D_GA := 5]
  dt.test2[, D_NLV := calc_nlv(B_LU_BRP, B_SOILTYPE_AGR, A_N_RT, A_CN_FR, D_OC, D_BDS, D_GA)]
    
  # plot output
  p2 <- ggplot(data = dt.test2,
               aes(x = A_N_RT, y = D_NLV)) + geom_line() + geom_point(size=3) +  theme_bw() + 
    theme(plot.title = element_text(size=10),
              legend.text = element_text(size=10),
              axis.text = element_text(size=10))+
    ylab('N supplying capacity (kg N / ha)') + xlab('Total Nitrogen content (mg / kg)') +
    ggtitle('Relationship NLV (grasland) and N total') 

  # plot side by side
  p1 + p2
  
```


The total N supply is evaluated via a parabolic scoring function with an optimum of 100 kg N / ha in arable fields and 140 kg N / ha in grassland fields.

<br />

### Phosphorus Availability Index
The phosphorus supply of an agricultural soil can be derived from methods measuring either a capacity or an intensity P pools in soil or a combination of both. The latter approach is followed for grassland and maize fields, where the optimum P supply is derived from multiple field experiments across the Netherlands. For arable farming systems, the phosphorus supply is derived from one measured P pool reflecting the P status as well as controlling the crop response to P inputs. The availability is highly controlled by chemical sorption and desorption equilibria, and affected by the iron and aluminum oxides content of soils as well as the availability of oxygen. A phosphorus supply around 4.8 (unitless index, for maize and grassland fields) and 45 (mg P2O5 per liter, for arable fields)  is optimal for crop production.

To estimate the P availability of the soil, one can make use of the function `calc_phosphate_availability`. The function requires the following input: the BRP crop code, and the concentration of P measured via extraction of ammonium lactate (A_P_AL), water (A_P_WA) or 0.01M CaCl2 (A_P_CC). 

For more details, see `?calc_phosphate_availability`. 


```{r, results = FALSE}

  # estimate phosphate availability index (unitless)
  dt[, D_PBI := calc_phosphate_availability(B_LU_BRP, A_P_AL, A_P_CC, A_P_WA)]

```


An example of this function is illustrated below for both an arable and a grassland field (left figure) as well for the relationship between P-CaCl2 and D_PBI for a grassland soil. In this case the P capacity is left constant (so, A_P_AL is not changing) whereas the bio-available P pool is increasing from 0.5 to 10 mg P / kg.

<br />

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # define an arable field with two landuses
  dt.test <- binnenveld[ID %in% c(1,3)]
  
  # estimate PBI
  dt.test[, D_PBI := calc_phosphate_availability(B_LU_BRP, A_P_AL, A_P_CC, A_P_WA)]
  
  # add group for figure
  dt.test[,luse := fifelse(ID==1,'arable','grasland')]
  
  # estimate averaged PBI for both soils
  dt.test1 <- dt.test[,list(D_PBI = mean(D_PBI)), by = 'luse']
  
   # plot output
  p1 <- ggplot(data = dt.test1,aes(y = D_PBI, x = luse, fill = luse)) + 
        geom_col(show.legend = FALSE) +  theme_bw() + 
        ylab('P availability index (-)') + xlab('Landuse') + ylim(0,7)+
        theme(plot.title = element_text(size=10),
              axis.text = element_text(size=10)) + 
        ggtitle('P availability index')
     
  
  # estimate PBI for grassland soil over range of P-CaCl2 levels in soil
  dt.test2 <- dt.test[ID==3]
  
  # overwrite P-CaCl2 levels
  dt.test2[, A_P_CC := seq(0.5,10,length.out = .N)]
  dt.test2[, B_SOILTYPE_AGR := 'dekzand']
  dt.test2[, D_PBI := calc_phosphate_availability(B_LU_BRP, A_P_AL, A_P_CC, A_P_WA)]
    
  # plot output
  p2 <- ggplot(data = dt.test2,
               aes(x = A_P_CC, y = D_PBI)) + geom_line() + geom_point(size=3) +  theme_bw() + 
        theme(plot.title = element_text(size=10),
                  legend.text = element_text(size=10),
                  axis.text = element_text(size=10))+
        ylab('P availability index (-)') + xlab('Available P-CaCl2 (mg / kg)') +
        ggtitle('Relationship PBI (grasland) and P-CaCl2') 

  # plot side by side
  p1 + p2
  
```


### Potassium supply
The potassium supply is strongly controlled by the distribution of K over the possible K pools in soil, being controlled by sorption and desorption processes, the cation exchange capacity of the soil and competing ions in soil solution and present at the exchange complex. Building upon agronomic research done in last decade, the K supply is quantified for grasslands using the plant available K fraction (measured via CaCl2-extraction, K-CaCl2) as well as the CEC. For maize it is quantified using K-CaCl2 and for arable fields its quantified from K-CaCl2 and the CEC while accounting for the clay content, the organic matter content and the pH of the soil. 

To estimate the K supplying capacity of the soil, one can make use of the function `calc_potassium_availability`. The function requires the following input: the BRP crop code, the soil type, the organic matter content, the clay content, the pH, the cation exchange capacity (CEC), the occupation of the CEC with potassium (%) and the directly available concentration of K as determined with a soil extraction with 0.01M CaCl2. 

For more details, see `?calc_potassium_availability`. 


```{r, results = FALSE}

  # estimate potassium availability index (unitless)
  dt[, D_K :=  calc_potassium_availability(B_LU_BRP, B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI, 
                                           A_PH_CC, A_CEC_CO, A_K_CO_PO, A_K_CC)]

```


An example of this function is illustrated below for both an arable and a grassland field (left figure) as well for the relationship between K-CaCl2 and D_K for a grassland soil. In this case the occupation of the CEC with K is left constant (so, A_K_CO_PO is not changing) whereas the bio-available K pool is increasing from 5 to 500 mg K / kg.

<br />

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # define an arable field with two landuses
  dt.test <- binnenveld[ID %in% c(1,3)]
  
  # estimate K-availability index
  dt.test[, D_K :=  calc_potassium_availability(B_LU_BRP, B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI, 
                                           A_PH_CC, A_CEC_CO, A_K_CO_PO, A_K_CC)]
  
  # add group for figure
  dt.test[,luse := fifelse(ID==1,'arable','grasland')]
  
  # estimate averaged K-availability indices for both soils
  dt.test1 <- dt.test[,list(D_K = mean(D_K)), by = 'luse']
  
   # plot output
  p1 <- ggplot(data = dt.test1,aes(y = D_K, x = luse, fill = luse)) + 
        geom_col(show.legend = FALSE) +  theme_bw() + 
        ylab('K availability index (-)') + xlab('Landuse') + 
        theme(plot.title = element_text(size=10),
              axis.text = element_text(size=10)) + 
        ggtitle('K availability index')
     
  
  # estimate K-supply for grassland soil over range of K-CaCl2 levels in soil
  dt.test2 <- dt.test[ID==3]
  
  # overwrite K-CaCl2 levels
  dt.test2[, A_K_CC := seq(5,500,length.out = .N)]
  dt.test2[, B_SOILTYPE_AGR := 'dekzand']
  dt.test2[, D_K :=  calc_potassium_availability(B_LU_BRP, B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI, 
                                           A_PH_CC, A_CEC_CO, A_K_CO_PO, A_K_CC)]
    
  # plot output
  p2 <- ggplot(data = dt.test2,
               aes(x = A_K_CC, y = D_K)) + geom_line() + geom_point(size=3) +  theme_bw() + 
        theme(plot.title = element_text(size=10),
                  legend.text = element_text(size=10),
                  axis.text = element_text(size=10))+
        ylab('K availability index (-)') + xlab('Available K-CaCl2 (mg / kg)') +
        ggtitle('Relationship K-availability (grasland) and K-CaCl2') 

  # plot side by side
  p1 + p2
  
```


### Sulpher supply
The amount of sulphur in the soil that is available for crop uptake is determined by the mineral sulphur supply remaining in the soil after winter (Smin), mineralisation of sulphur in the soil during the growing season (sulphur supply capacity), sulphur deposition, sulphur supply by irrigation during cultivation, capillary rise of sulphur-containing groundwater and by possible losses. Sulphur supply, capillary rise and Smin are quantitatively the most important inputs of sulphur. However, all three can vary considerably. The contribution from sulphur deposition is only minor throughout the Netherlands. A substantial amount of sulphur can also be supplied through irrigation with sulphurous spring water. This supply, however, varies greatly, depending on the amount sprayed and the sulphur content of the water.

To estimate the S supplying capacity of the soil, one can make use of the function `calc_slv`. The function requires the following input: the BRP crop code, the soil type, the agricultural region within the Netherlands (used to estimate default S deposition rates), the organic matter content, the total S content of the soil and the bulk density. 

For more details, see `?calc_slv`. 


```{r, results = FALSE}

  # estimate S supplying capacity
  dt[, D_SLV := calc_slv(B_LU_BRP, B_SOILTYPE_AGR, B_AER_CBS,A_SOM_LOI,A_S_RT, D_BDS)]

```


An example of this function is illustrated below for both an arable and a grassland field (left figure) as well for the relationship between total S and D_SLV for a grassland soil. In this case all soil properties driving S supplying capacity are left constant whereas the total S pool is increasing from 500 to 5000 mg S / kg.


<br />

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # define an arable field with two landuses
  dt.test <- binnenveld[ID %in% c(1,3)]
  
  # estimate bulk density
  dt.test[, D_BDS := calc_bulk_density(B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI)]

  # estimate Mg-availability index
  dt.test[,D_SLV := calc_slv(B_LU_BRP, B_SOILTYPE_AGR, B_AER_CBS,A_SOM_LOI,A_S_RT, D_BDS)]
  
  # add group for figure
  dt.test[,luse := fifelse(ID==1,'arable','grasland')]
  
  # estimate averaged SLV for both soils
  dt.test1 <- dt.test[,list(D_SLV = mean(D_SLV)), by = 'luse']
  
   # plot output
  p1 <- ggplot(data = dt.test1,aes(y = D_SLV, x = luse, fill = luse)) + 
        geom_col(show.legend = FALSE) +  theme_bw() + 
        ylab('S supplying capacity (kg S / ha)') + xlab('Landuse') + 
        theme(plot.title = element_text(size=10),
              axis.text = element_text(size=10)) + 
        ggtitle('S supplying capacity (kg S / ha)')
     
  
  # estimate SLV for grassland soil over range of total S levels in soil
  dt.test2 <- dt.test[ID==3]
  
  # overwrite A_S_RT levels
  dt.test2[, A_S_RT := seq(500,5000,length.out = .N)]
  dt.test2[, B_SOILTYPE_AGR := 'dekzand']
  dt.test2[, D_SLV := calc_slv(B_LU_BRP, B_SOILTYPE_AGR, B_AER_CBS,A_SOM_LOI,A_S_RT, D_BDS)]
    
  # plot output
  p2 <- ggplot(data = dt.test2,
               aes(x = A_S_RT, y = D_SLV)) + geom_line() + geom_point(size=3) +  theme_bw() + 
        theme(plot.title = element_text(size=10),
                  legend.text = element_text(size=10),
                  axis.text = element_text(size=10))+
        ylab('S supplying capacity (kg S / ha)') + xlab('total S (mg / kg)') +
        ggtitle('Relationship SLV (grasland) and total S') 

  # plot side by side
  p1 + p2
  
```



### Magnesium supply
Deficiencies for magnesium often occur on sandy soils with low organic matter and low pH. The supply of Mg for arable fields as well as grassland fields on sandy soils is directly related to the plant available Mg-content (Mg-CaCl2). For grassland fields on clay and peat soils the Mg-supply is indirectly quantified given the required composition of the grass (for intake by cows) and depends on the CEC, the organic matter and clay content, pH, as well as the K-availability assessed by K-CaCl2 and K-CEC. 

To estimate the Mg supplying capacity of the soil, one can make use of the function `calc_magnesium_availability`. The function requires the following input: the BRP crop code, the soil type, the organic matter content, the clay content, the pH, the cation exchange capacity (CEC), the occupation of the CEC with potassium (%) and the directly available concentration of both Mg and K as determined with a soil extraction with 0.01M CaCl2. 

For more details, see `?calc_magnesium_availability`. 


```{r, results = FALSE}

  # estimate magnesium availability index (unitless)
  dt[, D_MG := calc_magnesium_availability(B_LU_BRP, B_SOILTYPE_AGR, A_SOM_LOI, 
                                           A_CLAY_MI, A_PH_CC, A_CEC_CO,
                                           A_K_CO_PO, A_MG_CC, A_K_CC)]

```


An example of this function is illustrated below for both an arable and a grassland field (left figure) as well for the relationship between Mg-CaCl2 and D_MG for a grassland soil. In this case all soil properties driving Mg availability in soil are left constant whereas the bio-available Mg pool is increasing from 5 to 500 mg Mg / kg.

<br />

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # define an arable field with two landuses
  dt.test <- binnenveld[ID %in% c(1,3)]
  
  # estimate Mg-availability index
  dt.test[, D_MG := calc_magnesium_availability(B_LU_BRP, B_SOILTYPE_AGR, A_SOM_LOI, 
                                           A_CLAY_MI, A_PH_CC, A_CEC_CO,
                                           A_K_CO_PO, A_MG_CC, A_K_CC)]
  
  # add group for figure
  dt.test[,luse := fifelse(ID==1,'arable','grasland')]
  
  # estimate averaged Mg-availability indices for both soils
  dt.test1 <- dt.test[,list(D_MG = mean(D_MG)), by = 'luse']
  
   # plot output
  p1 <- ggplot(data = dt.test1,aes(y = D_MG, x = luse, fill = luse)) + 
        geom_col(show.legend = FALSE) +  theme_bw() + 
        ylab('Mg availability index (-)') + xlab('Landuse') + 
        theme(plot.title = element_text(size=10),
              axis.text = element_text(size=10)) + 
        ggtitle('Mg availability index')
     
  
  # estimate Mg-supply for grassland soil over range of Mg-CaCl2 levels in soil
  dt.test2 <- dt.test[ID==3]
  
  # overwrite Mg-CaCl2 levels
  dt.test2[, A_MG_CC := seq(5,500,length.out = .N)]
  dt.test2[, B_SOILTYPE_AGR := 'dekzand']
  dt.test2[, D_MG := calc_magnesium_availability(B_LU_BRP, B_SOILTYPE_AGR, A_SOM_LOI, 
                                           A_CLAY_MI, A_PH_CC, A_CEC_CO,
                                           A_K_CO_PO, A_MG_CC, A_K_CC)]
    
  # plot output
  p2 <- ggplot(data = dt.test2,
               aes(x = A_MG_CC, y = D_MG)) + geom_line() + geom_point(size=3) +  theme_bw() + 
        theme(plot.title = element_text(size=10),
                  legend.text = element_text(size=10),
                  axis.text = element_text(size=10))+
        ylab('Mg availability index (-)') + xlab('Available Mg-CaCl2 (mg / kg)') +
        ggtitle('Relationship Mg-availability (grasland) and Mg-CaCl2') 

  # plot side by side
  p1 + p2
  
```


### Copper supply
Copper (Cu) occurs in soil solution largely in a complexed form, mainly with dissolved organic matter. Root uptake of Cu probably occurs as free Cu ions (Cu2+). The availability of complexed Cu for uptake is lower than that of Cu2+ and depends on the stability of the complex. Complexation of Cu, however, may increase the mobility of Cu in soils. Besides complexation of Cu to dissolved or solid organic matter, Cu can be adsorbed to Fe-, Al- and Mn-(hydr)oxides and clay minerals. Cu desorbed from the mentioned soil components is available for uptake. The degree of adsorption increases at higher pH, which decreases the availability. Other factors that influence Cu availability are nitrogen and phosphate content. Nitrogen and Cu show positive interactions. Decreasing N use will lead to lower levels in crops. Phosphate can drive Cu away from the adsorption sites and this leads to Cu deficiency in the long run.

The supply and availability of Cu depends on the plant available Cu-pool in soil (Cu-CaCl2) or the reactive Cu pool (Cu-HNO3) for both arable and grassland soils. The optimum value is derived from historical field experiments and is established on a level of 5 to 10 mg / kg. The risk of Cu deficiency can be calculated based on the Cu content in the soil and relevant soil characteristics that influence its availability. The following measurements are used to calculate Cu plant availability: land use, Cu-CaCl2, clay content, the organic matter content, and the plant extractable levels of K and Mn. The last soil properties are used for an empirical function to estimate Cu-HNO3 (the basis for fertilizer recommendations based on long-term field experiments) from Cu-CaCl2 (being currently measured in agricultural laboratories). In OBIC, one can make use of the function `calc_copper_availability`. The function requires the input mentioned above. 

For more details, see `?calc_copper_availability`. 



```{r, results = FALSE}

  # estimate Cu availability index
  dt[, D_CU := calc_copper_availability(B_LU_BRP, A_SOM_LOI, A_CLAY_MI, A_K_CC, A_MN_CC, A_CU_CC)]

```


An example of this function is illustrated below for both an arable and a grassland field (left figure) as well for the relationship between Cu-CaCl2 and D_CU for a grassland soil. In this case all soil properties driving Cu availability in soil are left constant whereas the bio-available Cu pool is increasing from 5 to 500 ug Cu / kg.

<br />

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # define an arable field with two landuses
  dt.test <- binnenveld[ID %in% c(1,3)]
  
  # estimate Cu-availability index
  dt.test[, D_CU := calc_copper_availability(B_LU_BRP, A_SOM_LOI, A_CLAY_MI, A_K_CC, A_MN_CC, A_CU_CC)]
  
  # add group for figure
  dt.test[,luse := fifelse(ID==1,'arable','grasland')]
  
  # estimate averaged Cu-availability indices for both soils
  dt.test1 <- dt.test[,list(D_CU = mean(D_CU)), by = 'luse']
  
   # plot output
  p1 <- ggplot(data = dt.test1,aes(y = D_CU, x = luse, fill = luse)) + 
        geom_col(show.legend = FALSE) +  theme_bw() + 
        ylab('Cu availability index (-)') + xlab('Landuse') + 
        theme(plot.title = element_text(size=10),
              axis.text = element_text(size=10)) + 
        ggtitle('Cu availability index')
     
  
  # estimate Cu-supply for grassland soil over range of Cu-CaCl2 levels in soil
  dt.test2 <- dt.test[ID==3]
  
  # overwrite Cu-CaCl2 levels
  dt.test2[, A_CU_CC := seq(5,500,length.out = .N)]
  dt.test2[, B_SOILTYPE_AGR := 'dekzand']
  dt.test2[, D_CU := calc_copper_availability(B_LU_BRP, A_SOM_LOI, A_CLAY_MI, A_K_CC, A_MN_CC, A_CU_CC)]
    
  # plot output
  p2 <- ggplot(data = dt.test2,
               aes(x = A_CU_CC, y = D_CU)) + geom_line() + geom_point(size=3) +  theme_bw() + 
        theme(plot.title = element_text(size=10),
                  legend.text = element_text(size=10),
                  axis.text = element_text(size=10))+
        ylab('Cu availability index (-)') + xlab('Available Cu-CaCl2 (ug / kg)') +
        ggtitle('Relationship Cu-availability (grasland) and Cu-CaCl2') 

  # plot side by side
  p1 + p2
  
```


### Zinc supply
The processes involved in the behavior of zinc (Zn) in soil are the same as for Cu. Besides Zn2+, Zn occurs in the soil solution in a complex form with dissolved organic matter. Binding of Zn to solid soil constituents occurs to clay minerals, metal (hydr)oxides and solid organic matter. The affinity of Zn for binding to these materials is lower than that of Cu. Zinc is taken up by plant roots as Zn2+. Because the concentration of Zn2+ in the soil solution is low and Zn is not very mobile, it is only transported over short distances. Therefore, a sufficiently large root system is very important for a good Zn uptake. The development of the root system is strongly influenced by the phosphate supply of the crop. In herbaceous crops, the secretion of phytosiderophores in response to a Fe or Zn deficiency strongly increases Zn uptake. These organic compounds complex Zn and increase Zn mobility in the soil, which facilitates Zn transport to the root and uptake as Zn2+. A high pH, cold and wet conditions and a lot of organic matter in the soil decrease the availability of Zn.

The supply of zinc depends on the plant availability of Zn such as determined via CaCl2 extraction method (Zn-CaCl2), while accounting for soil type, land use and pH.Critical threshold values for Zn has been derived from optimum Zn contents of the growing crop. In OBIC, one can make use of the function `calc_zinc_availability`. The function requires the following input: land use, soil type, pH and the bioavailable Zn-content as measured via n extraction with 0.01M CaCl2. 

For more details, see `?calc_zinc_availability`. 



```{r, results = FALSE}

  # estimate Zn availability index
  dt[,  D_ZN := calc_zinc_availability(B_LU_BRP, B_SOILTYPE_AGR, A_PH_CC, A_ZN_CC)]

```


An example of this function is illustrated below for both an arable and a grassland field (left figure) as well for the relationship between Zn-CaCl2 and D_ZN for a grassland soil. In this case all soil properties driving Zn availability in soil are left constant whereas the bio-available Zn pool is increasing from 50 to 5000 ug Zn / kg.

<br />

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # define an arable field with two landuses
  dt.test <- binnenveld[ID %in% c(1,3)]
  
  # estimate Zn-availability index
  dt.test[, D_ZN := calc_zinc_availability(B_LU_BRP, B_SOILTYPE_AGR, A_PH_CC, A_ZN_CC)]
  
  # add group for figure
  dt.test[,luse := fifelse(ID==1,'arable','grasland')]
  
  # estimate averaged Zn-availability indices for both soils
  dt.test1 <- dt.test[,list(D_ZN = mean(D_ZN)), by = 'luse']
  
   # plot output
  p1 <- ggplot(data = dt.test1,aes(y = D_ZN, x = luse, fill = luse)) + 
        geom_col(show.legend = FALSE) +  theme_bw() + 
        ylab('Zn availability index (-)') + xlab('Landuse') + 
        theme(plot.title = element_text(size=10),
              axis.text = element_text(size=10)) + 
        ggtitle('Zn availability index')
     
  
  # estimate Zn-supply for grassland soil over range of Zn-CaCl2 levels in soil
  dt.test2 <- dt.test[ID==3]
  
  # overwrite Zn-CaCl2 levels
  dt.test2[, A_ZN_CC := seq(50,5000,length.out = .N)]
  dt.test2[, B_SOILTYPE_AGR := 'dekzand']
  dt.test2[, D_ZN := calc_zinc_availability(B_LU_BRP, B_SOILTYPE_AGR, A_PH_CC, A_ZN_CC)]
    
  # plot output
  p2 <- ggplot(data = dt.test2,
               aes(x = A_ZN_CC, y = D_ZN)) + geom_line() + geom_point(size=3) +  theme_bw() + 
        theme(plot.title = element_text(size=10),
                  legend.text = element_text(size=10),
                  axis.text = element_text(size=10))+
        ylab('Zn availability index (-)') + xlab('Available Zn-CaCl2 (ug / kg)') +
        ggtitle('Relationship Zn-availability (grasland) and Zn-CaCl2') 

  # plot side by side
  p1 + p2
  
```


### Soil Acidity
The pH is a measure of the acidity of the soil. The pH influences the soil quality and crop growth, among others through availability of nutrients and (heavy) metals and activity of soil life. In addition, the pH influences also various soil functions such as nutrient buffering and aggregate stability. The optimum soil pH for crop growth differs per crop. Potatoes prefer a lower pH and cereals, maize, sugar beet and field vegetables a higher pH. Organic matter reduces the negative effect of pH on crop growth. On sandy soils with an organic matter content of 15%, the desired pH can be up to half a unit lower than on low organic matter content soils. On clay soils the desired pH is higher than on sandy soils. With a clay percentage above 10, the desired pH is half a unit higher than with low clay percentages.

The optimum pH is determined in historical field experiments given crop rotation plan, soil texture, soil organic matter and initial pH. The derivation of the optimum pH can be estimated via the OBIC using the function `calc_ph_delta`. This function estimates how much the pH has to increase by liming to be optimum for crop development. The original empirical relationships are based on pH measurements done in KCl-extraction. Since the current pH analysis methods all make use of 0.01M CaCl2, this pH value can easily be transformed to pH-KCl given an empirical relationship between both. 

For more details, see `?calc_ph_delta`. 


```{r, results = FALSE, eval = FALSE}

  # estimate distance to required pH
  dt[, D_PH_DELTA := calc_ph_delta(B_LU_BRP, B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI, A_PH_CC, D_CP_STARCH,
                                   D_CP_POTATO, D_CP_SUGARBEET, D_CP_GRASS, D_CP_MAIS, D_CP_OTHER)]

```


An example of this function is illustrated below for both an arable and a grassland field (left figure) as well for the relationship between the initial and desired pH change for a grassland soil. In this case all soil properties driving pH in soil are left constant whereas the initial pH is increasing from 3 to 10.

<br />

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # define an arable field with two landuses
  dt.test <- binnenveld[ID %in% c(1,3)]
  
  # Calculate the crop rotation fraction
  dt.test[, D_CP_STARCH := calc_rotation_fraction(ID, B_LU_BRP, crop = "starch")]
  dt.test[, D_CP_POTATO := calc_rotation_fraction(ID, B_LU_BRP, crop = "potato")]
  dt.test[, D_CP_SUGARBEET := calc_rotation_fraction(ID, B_LU_BRP, crop = "sugarbeet")]
  dt.test[, D_CP_GRASS := calc_rotation_fraction(ID, B_LU_BRP, crop = "grass")]
  dt.test[, D_CP_MAIS := calc_rotation_fraction(ID, B_LU_BRP, crop = "mais")]
  dt.test[, D_CP_OTHER := calc_rotation_fraction(ID, B_LU_BRP, crop = "other")]
    
  # estimate delta-pH
  dt.test[, D_PH_DELTA := calc_ph_delta(B_LU_BRP, B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI, A_PH_CC, D_CP_STARCH,
                                   D_CP_POTATO, D_CP_SUGARBEET, D_CP_GRASS, D_CP_MAIS, D_CP_OTHER)]
  
  # add group for figure
  dt.test[,luse := fifelse(ID==1,'arable','grasland')]
  
  # estimate averaged delta-pH for both soils
  dt.test1 <- dt.test[,list(D_PH_DELTA = mean(D_PH_DELTA)), by = 'luse']
  
   # plot output
  p1 <- ggplot(data = dt.test1,aes(y = D_PH_DELTA, x = luse, fill = luse)) + 
        geom_col(show.legend = FALSE) +  theme_bw() + 
        ylab('desired pH-change (-)') + xlab('Landuse') + 
        theme(plot.title = element_text(size=10),
              axis.text = element_text(size=10)) + 
        ggtitle('desired pH-change (-)')
     
  
  # estimate desired pH-change for grassland soil over range of pH levels in soil
  dt.test2 <- dt.test[ID==3]
  
  # overwrite initial pH
  dt.test2[, A_PH_CC := seq(3,10,length.out = .N)]
  dt.test2[, B_SOILTYPE_AGR := 'dekzand']
  dt.test2[, D_PH_DELTA := calc_ph_delta(B_LU_BRP, B_SOILTYPE_AGR, A_SOM_LOI, A_CLAY_MI, A_PH_CC, D_CP_STARCH,
                                   D_CP_POTATO, D_CP_SUGARBEET, D_CP_GRASS, D_CP_MAIS, D_CP_OTHER)]
    
  # plot output
  p2 <- ggplot(data = dt.test2,
               aes(y = D_PH_DELTA, x = A_PH_CC)) + geom_line() + geom_point(size=3) +  theme_bw() + 
        theme(plot.title = element_text(size=10),
                  legend.text = element_text(size=10),
                  axis.text = element_text(size=10))+
        ylab('desired pH-change (-)') + xlab('pH value') +
        ggtitle('Relationship desired pH-change (grasland) and pH') 

  # plot side by side
  p1 + p2
  
```



### Soil Cation Buffering
For good crop growth, sufficient nutrients must be available. Besides the actual nutrient availability in soil solution, the capacity of a soil to buffer nutrients is important to identify potential bottlenecks for crop production. Fertilization is needed when the buffer capacity or speed of replenishment is insufficient. Soils can vary greatly in their buffer capacity.  To supply a crop with cations (K, Mg, Na, Ca etc.), the buffer capacity is strongly determined by the cation exchange complex (CEC). Cations are adsorbed to negatively charged soil particles (clay, organic matter and oxides) and these cations may become available through exchange. For example, a growing plant extracts K+ from the soil solution. The adsorption complex will start to resupply because the amount of K+ in the soil solution has decreased. Once again, a balance is established between the K+ concentration in the soil solution and the quantity of K+ adsorbed. In practice, soil with a higher CEC are usually more fertilize and there is a lower risk of cation deficiencies during growing season. A general rule of thumb for the optimum CEC in soil has been established around a CEC value of about 100 mmol+/kg. CEC levels below 30-40 mmol+/kg are characterized by substantial crop limitation whereas values above 100 has no positive impact any more. The scientific underpinning of these guidelines is still poor. In OBIC the soil fertility is positively assessed in relation to the CEC, whereas this can be evaluated with the OBIC function `calc_cec`.

For more details, see `?calc_cec`. 



```{r, results = FALSE, eval = FALSE}

  # estimate importance of CEC supporting crop development
  dt[, D_CEC := calc_cec(A_CEC_CO)]

```



## Calculate series of physical soil functions
The open soil index quantifies and evaluates a series of physical soil functions related to the supply of water and the rootability of the top and subsoil. The different soil physical functions are illustrated and explained below. Soil physical functions include the capacity of soils to retain and supply, to resist wind erosion, topsoil sealing, subsoil compaction, and to deliver bearing capacity for machinal activities 

### Water Stress due to drought or wetness
The supply of water is of uttermost importance for crop development. The water supply depends on weather conditions and the capacity of soils to buffer and supply water during the growing season. Using long-term field experiments for multiple crops, empirical relationships between soil texture, ground water levels and yield depressions due to drought or too wet conditions have been developed (Van Bakel, 2005; Huinink, 2018). Within OBIC the crop yield response to droughtstress or wetnessstress can be estimated via the function `calc_waterstressindex`, requiring as input a soil code (representing unique combinations of soil type and geohydrological conditions controlling water availability), crop type, and groundwater level.

For more details, see `?calc_waterstressindex`. 


```{r, results = FALSE, eval = FALSE}

  # estimate risk on yield reduction to droughtstress or wetnessstress
  dt[, D_WSI_DS := calc_waterstressindex(B_HELP_WENR, B_LU_BRP, B_GWL_CLASS, WSI = 'droughtstress')]
  dt[, D_WSI_WS := calc_waterstressindex(B_HELP_WENR, B_LU_BRP, B_GWL_CLASS, WSI = 'wetnessstress')]

```


An example of this function is illustrated below for both an sandy and a clay soil given variation in the groundwater levels. In the left figure the yield reduction to drought stress (D_WSI_DS, red) and wetness stres (D_WSI_WS, green) are estimated for five fields with different soil types and crops. In the right figure the impact of variable groundwater dynamics are estimated for a single field.

<br />


```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # subset a serie of agricultural fields
  dt.test <- binnenveld[ID %in% c(1,2,3,4,5)]
  
  # estimate delta-pH
  dt.test[, D_WSI_DS := calc_waterstressindex(B_HELP_WENR, B_LU_BRP, B_GWL_CLASS, WSI = 'droughtstress')]
  dt.test[, D_WSI_WS := calc_waterstressindex(B_HELP_WENR, B_LU_BRP, B_GWL_CLASS, WSI = 'wetnessstress')]
  
  # estimate averaged moisture stress for the different fields
  dt.test1 <- dt.test[,list(D_WSI_DS = mean(D_WSI_DS),
                            D_WSI_WS = mean(D_WSI_WS)), by = 'ID']
  
  # melt dt.test1
  dt.test1 <- melt(dt.test1,id.vars = 'ID', variable.name = 'stresstype')
  
   # plot output
  p1 <- ggplot(data = dt.test1,aes(y = value, x = ID, fill = stresstype)) + 
        geom_bar(stat='identity') +  theme_bw() + ylim(0,50)+
        theme(legend.position = c(0.3,0.8),
              plot.title = element_text(size=10),
              legend.text = element_text(size=10)) +
        ylab('yield reduction (%)') + xlab('Field') + 
        ggtitle('Soil water stress index')
     
  
  # estimate moisture stress with variable Gt for an arable field
  dt.test2a <- dt.test[ID==1]
  dt.test2b <- copy(dt.test2a)[,B_GWL_CLASS := 'GtIII']
  dt.test2c <- copy(dt.test2a)[,B_GWL_CLASS := 'GtIV']
  dt.test2d <- copy(dt.test2a)[,B_GWL_CLASS := 'GtV']
  dt.test2e <- copy(dt.test2a)[,B_GWL_CLASS := 'GtVI']
  dt.test2 <- rbind(dt.test2a,dt.test2b,dt.test2c,dt.test2d,dt.test2e)
  
  # update WSI
  dt.test2[, D_WSI_DS := calc_waterstressindex(B_HELP_WENR, B_LU_BRP, B_GWL_CLASS, WSI = 'droughtstress')]
  dt.test2[, D_WSI_WS := calc_waterstressindex(B_HELP_WENR, B_LU_BRP, B_GWL_CLASS, WSI = 'wetnessstress')]
  
   # estimate averaged moisture stress for the different fields
  dt.test2 <- dt.test2[,list(D_WSI_DS = mean(D_WSI_DS),
                            D_WSI_WS = mean(D_WSI_WS)), by = 'B_GWL_CLASS']
  
  # melt dt.test2
  dt.test2 <- melt(dt.test2,id.vars = 'B_GWL_CLASS', variable.name = 'stresstype')
    
  # plot output
  p2 <- ggplot(data = dt.test2,aes(y = value, x = B_GWL_CLASS, fill = stresstype)) + 
        geom_bar(stat='identity') +  theme_bw() + ylim(0,30)+
        theme(legend.position = c(0.3,0.8),
              plot.title = element_text(size=10),
              legend.text = element_text(size=10)) +
        ylab('yield reduction (%)') + xlab('GWL_CLASS') + 
        ggtitle('Soil water stress index')

  # plot side by side
  p1 + p2
  
```

### Sealing Soil Surface and Wind Erodibility
The vulnerability of soils to sealing and wind erosion is derived from those properties controlling the susceptibility for both risks. These include the clay content as well as the organic matter content, where the vulnerability for sealing is derived from field expert-based observations and associated yield depression (Huinink, 2018; Heeres en Van Erp, 1999) in particular for some arable crops. The presence of soil sealing can be estimated via the function `calc_sealing_risk` whereas the risk for soil erosion due to wind is estimated via the function `calc_winderodibility`. The wind erodibility can be estimated from the inputs land use, clay and silt content of the topsoil whereas the risk for sealing depends on the clay content as well as the organic matter content of the soil.

Wind erosion is highly controlled by the mineralogical composition and texture as well as the presence of growing crops. Using laboratory experiments from wind tunnels, a exponential function has been developed to estimate the erodibility risk for soils, given the clay and silt content of a field. Grasslands are not susceptible for these two risks.

For more details, see `?calc_sealing_risk` and `?calc_winderodibility`. 


```{r, results = FALSE, eval = FALSE}

  # estimate the presence / risk for surface sealing and wind erodibility
  dt[, D_SE := calc_sealing_risk(A_SOM_LOI, A_CLAY_MI)]
  dt[, D_WE := calc_winderodibility(B_LU_BRP, A_CLAY_MI, A_SILT_MI)]

```


An example of these functions is illustrated below for an arable soil with variable clay content for both an cropland (with organic matter levels of 3.5%) and a grassland soil (with organic matter levels of 5%). Soil sealing is a potential risk and this risk is evaluated in the indicator function (explained and illustrated later) given the land use present. The wind erodibility is declining with the clay content and is not relevant for grassland due to the permanent crop cover. 

<br />


```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # make a data.table for sandy and clay soil with potatoes
  dt.test1 <- data.table(B_LU_BRP = rep(859,20),
                        A_CLAY_MI = c(seq(1,25,length.out = 10), seq(1,25,length.out = 10)),
                        A_SOM_LOI = c(rep(3.5,10),rep(3.5,10)),
                        A_SILT_MI = c(rep(5,10),rep(5,10)))

  # make a data.table for sandy and clay soil with grasland
  dt.test2 <- data.table(B_LU_BRP = rep(265,20),
                        A_CLAY_MI = c(seq(1,25,length.out = 10), seq(1,25,length.out = 10)),
                        A_SOM_LOI = c(rep(5,10),rep(5,10)),
                        A_SILT_MI = c(rep(5,10),rep(5,10)))
  
  # combine both
  dt.test <- rbind(dt.test1,dt.test2)
  dt.test[, B_SOILTYPE_AGR := fifelse(A_CLAY_MI < 15,'dekzand','rivierklei')]
  
  # estimate the presence / risk for surface sealing and wind erodibility
  dt.test[, D_SE := calc_sealing_risk(A_SOM_LOI, A_CLAY_MI)]
  dt.test[, D_WE := calc_winderodibility(B_LU_BRP, A_CLAY_MI, A_SILT_MI)]
  
  # set land use
  dt.test[,luse := fifelse(B_LU_BRP==265,'grass','cropland')]
   
  # plot output sealing
  p1 <- ggplot(data = dt.test,
               aes(x = A_CLAY_MI, y = D_SE, color = luse,group = luse)) + 
        geom_line() + geom_point(size=3) +  theme_bw() + 
        ylab('Soil Sealing (-)') + xlab('Clay content (%)') + ylim(4,12)+
        theme(legend.position = c(0.4,0.8),
              plot.title = element_text(size=10),
              legend.text = element_text(size=10)) + 
    ggtitle('Soil sealing given SOM and clay content')   

  # plot output wind erodibility
  p2 <- ggplot(data = dt.test,
               aes(x = A_CLAY_MI, y = D_WE, color = luse)) + 
        geom_line() + geom_point(size=3) +  theme_bw() + 
        ylab('Wind Erodibility (-)') + xlab('Clay content (%)') + ylim(0,1)+
        theme(legend.position = c(0.7,0.8),
              plot.title = element_text(size=10),
              legend.text = element_text(size=10)) + 
    ggtitle('Wind erodibility given soil texture and landuse') 
  
  # plot side by side
  p1 + p2
  
```


### Aggregate Stability and Crumbleability
Similarly to the vulnerability to sealing and wind erodiblity, the crumbleability of a soil (a measure for aggregate stability) as well as the potential risk for yield depressions (crop specific) can be derived from the clay content, the organic matter content and the pH (Huinink, 2011). Aggregate stability is additionally evaluated given the cation occupation of the cation exchange capacity. Clayey soils with 80% Ca, 8% Mg and about 3.5% K are recognized as soils with more stable aggregates due to the influence of electrostatic bindings of divalent cations. Within OBIC these two soil functions are calculated using `calc_crumbleability` and `calc_aggregatestability`.

For more details, see `?calc_crumbleability` and `?calc_aggregatestability`. 


```{r, results = FALSE, eval = FALSE}

  # assess the crumbleability and aggregate stability
  dt[, D_CR := calc_crumbleability(A_SOM_LOI, A_CLAY_MI,A_PH_CC)]
  dt[, D_AS := calc_aggregatestability(B_SOILTYPE_AGR,A_SOM_LOI,A_K_CO_PO,A_CA_CO_PO,A_MG_CO_PO)]
 

```



### Subsoil compaction
The risk of subsoil compaction for each field was assessed using soil texture, land use and a soil compaction model SOCOMO. The model determines whether the usual wheel loads for that land use exceed the strength of the subsurface in wet or humid conditions. Subsequently it was determined on the basis of the soil properties and groundwater levels whether the subsurface is extra sensitive to compaction or whether natural recovery is possible due to, for example, drought shrinkage. Strength, land use and soil properties together determine the risk of subsoil compaction. The presence of subsoil compaction is derived from a national map with the estimated risk. As such, there is not calculation used. Recent field experimental data support the accuracy of this risk assessment map.

Within the OBIC, this estimate can be adapted to local site specific insights derived from the Visual Soil Assessment Methodology. Visual soil assessments (VSA) asses soil quality in the field, by digging a hole and assess several soil quality indicators visually. Soil quality indicators are grass cover, number of roots, number of biopores, soil color, soil structure, soil compaction and number of gray spots. Visual soil assessments are more and more used by farmers in the Netherlands, but not specifically developed for various soil types. 

Below is illustrated how the obic_field wrapper deals with the parameters derived from the VSA methodology. When scores are present for the presence of earthworms (A_EW_BCS), the occurence of subsurface compaction (A_SC_BCS), limited root development (A_RD_BCS), ponding in the spring (wet conditions after rainfall, A_P_BCS) and clear presence of visible tracks / rutting or trampling on the land (A_RT_BCS), then the original risk assessment map for subsoil compaction is overwritten the score for the relevant VSA indicators. Similarly, the soil aggregate stability is adapted when VSA indicators are available for earthworm density (A_EW_BCS), soil structure (A_SS_BCS) and presence of cracks in the topsoil (A_C_BCS).

```{r, results = FALSE, eval = FALSE}

    # overwrite soil physical functions for compaction when BCS is available
    dt[,D_P_CO := (3 * A_EW_BCS + 3 * A_SC_BCS + 3 * A_RD_BCS  - 2 * A_P_BCS - A_RT_BCS)/18]
    dt[,D_P_CO := pmax(0, D_P_CO)]
    dt[,I_P_CO := fifelse(is.na(D_P_CO),I_P_CO,D_P_CO)]
    
    # overwrite soil physical functions for aggregate stability when BCS is available
    dt[,D_P_CEC := (3 * A_EW_BCS + 3 * A_SS_BCS - A_C_BCS)/12]
    dt[,D_P_CEC := pmax(0, D_P_CEC)]
    dt[,I_P_CEC := fifelse(is.na(D_P_CEC),I_P_CEC,D_P_CEC)]
```


### Water retention in topsoil
The review study of various soil health instruments (Buneman et al., 2016) identifies a variety of indicators that can be used to gain more insight into the water retaining capacity of a soil. Frequently used indicators are the amount of plant-available water (14 x), the water holding capacity (11 x), the moisture content at field capacity (11 x) and the saturated permeability (10 x). These indicators can be measured or estimated indirectly via so-called pedotransfer functions (Wosten et al., 2001).  

The OBIC includes functions to derive the following water-related indicators in function `calc_waterretention`:

* the moisture content at wilting point and field capacity
* The amount of plant available water (mm) 
* the flow limit or porosity, also called water holding capacity, and
* the saturated permeability.

The basic indicator used in the OBIC is the amount of plant available water. The capacity of the soil to buffer water during the growing season is estimated from pedotransferfunctions (pdf) for plant available water in the topsoil. The default pdf is derived from a European database, and enables one to estimate the van Genuchten parameters for the pF-curve. Other options are also possible.

For more details, see `?calc_waterretention`. 


```{r, results = FALSE, eval = FALSE}

  # estimate the plant available water in topsoil
  dt[, D_WRI := calc_waterretention(A_CLAY_MI,A_SAND_MI,A_SILT_MI,A_SOM_LOI,type = 'plant available water')]
  
  # estimate the water holding capacity in topsoil
  dt[, D_WRI := calc_waterretention(A_CLAY_MI,A_SAND_MI,A_SILT_MI,A_SOM_LOI,type = 'water holding capacity')]
  
  # estimate the moisture content of the wilting point in topsoil
  dt[, D_WRI := calc_waterretention(A_CLAY_MI,A_SAND_MI,A_SILT_MI,A_SOM_LOI,type = 'wilting point')]
  
  # estimate the moisture content of the field capacity in topsoil
  dt[, D_WRI := calc_waterretention(A_CLAY_MI,A_SAND_MI,A_SILT_MI,A_SOM_LOI,type = 'field capacity')]
  
  # estimate the saturated permeability in topsoil
  dt[, D_WRI := calc_waterretention(A_CLAY_MI,A_SAND_MI,A_SILT_MI,A_SOM_LOI,type = 'Ksat')]

```



## Calculate series of biological soil functions
The open soil index quantifies and evaluates a series of soil biological functions related to the composition of the microbiome, suppression of soil borne pests and plagues and plant parasitic nematodes.

### Potentially mineralizable nitrogen
The activity of soil fungi and bacteria has been measured via incubation trials for decades. In such incubation trials, the decomposition of soil organic nitrogen is measured over a period of time. The amount of nitrogen that mineralises in such trials is called the 'potential mineralisable nitrogen'(PMN). This indicates the ability of the soil to provide a crop with readily available nitrogen. In the Netherlands, this parameter is also used as an indicator for activity of soil microbes. This is possible because during the incubation, a soil sample is submerged in water, creating an anerobic enviroment, killing aerobic bacteria and fungi, which, together with some labile organic matter, are decomposed by anerobic soil microbes.



For more information see `?calc_pmn`.

```{r, results = FALSE, eval = FALSE}

  # calculate the index for the potential mineralisable nitrogen pool
  dt[, D_PMN := calc_pmn(B_LU_BRP, B_SOILTYPE_AGR, A_N_PMN)]


```

The PMN measured by routine laboratories in the Netherlands may be overestimated. Therefore, a landusedependent correction factor is used. The PMN index is capped at 500. Below an example of this function is given.

```{r,fig.width = 7, fig.height = 4,fig.fullwidth = TRUE,echo=FALSE}

  # define test data table with increasing PMN and different land uses
  dt.test <- data.table(B_LU_BRP = c(rep(265,12), rep(234,12)), B_SOILTYPE_AGR = 'rivierklei',A_N_PMN = rep(seq(0,550,50),2), A_N_PMN2 = rep(seq(0,220,20),2))
  dt.test2 <- data.table(B_LU_BRP = c(rep(265,12), rep(234,12)), B_SOILTYPE_AGR = 'dekzand',A_N_PMN = rep(seq(0,550,50),2), A_N_PMN2 = rep(seq(0,220,20),2))
  dt.test <- rbindlist(list(dt.test, dt.test2))
  
  # Add unique identifyer
  id <- as.data.table(unique(dt.test[,.(B_LU_BRP, B_SOILTYPE_AGR)]))
  id <- id[,landuse := fifelse(B_LU_BRP == 265 & B_SOILTYPE_AGR == 'rivierklei', 'grass or maize on clay', fifelse(
    B_LU_BRP == 265 & B_SOILTYPE_AGR == 'dekzand', 'grass or maize on sand', fifelse(
      B_LU_BRP == 234 & B_SOILTYPE_AGR == 'rivierklei', 'arable on clay', 'arable on sand')
  ))]
  dt.test <- merge.data.table(dt.test, id, by = c('B_LU_BRP', 'B_SOILTYPE_AGR'))
  
  # Calculate PMN index
  dt.test <- dt.test[,D_PMN := calc_pmn(B_LU_BRP, B_SOILTYPE_AGR, A_N_PMN)]

  # plot output
  p1 <- ggplot(data = dt.test,
               aes(y = D_PMN, x = A_N_PMN, group = landuse, colour = landuse)) + geom_line() + geom_point(size=3) +  theme_bw() + 
        theme(plot.title = element_text(size=10),
                  legend.text = element_text(size=10),
                  axis.text = element_text(size=10),
              legend.position = c(0.75,0.15))+
        ylab('PMN index') + xlab('Microbial actitivy') +
        ggtitle('Relationship between measured PMN and\n adjusted PMN index') 

  # calculate indicator with A_PMN2 (which has lower numbers)
  dt.test <- dt.test[, I_PMN2 := ind_pmn(calc_pmn(B_LU_BRP, B_SOILTYPE_AGR, A_N_PMN2))]
  
  # Plot indicator
    p2 <- ggplot(data = dt.test,
               aes(y = I_PMN2, x = A_N_PMN2, group = landuse, colour = landuse)) + geom_line() + geom_point(size=3) +  theme_bw() + 
        theme(plot.title = element_text(size=10),
                  legend.text = element_text(size=10),
                  axis.text = element_text(size=10),
              legend.position = c(0.75,0.15))+
        ylab('PMN indicator score') + xlab('Microbial actitivy') +
        ggtitle('Relationship between measured PMN and PMN\n indicator score') 
    
  # plot side by side
  p <- ggarrange(p1, p2, ncol = 2, common.legend = TRUE)
  p
  
```

### Disease resistance

### Nematodes



### Earthworm density (from VSA)


## Calculate series of environmental soil functions
The open soil index quantifies and evaluates a series of environmental soil functions related to xxxx. Soil environmental functions include xxxx.

### N retention preventing N leaching

### N retention preventing N runoff




## Evaluate Soil Management
Add text of LDBB.

### Measures

### Crop rotation and SOM balance

### Recommendations



## Integrating Soil Functions into OBIC scores

### Crop and soil dependent applicability of functions

### Integrating over time

### Aggregated soil functions

### Holistic OBI score






